The error when inferring phylogenies with incipient species by a birth-death model

### 2018-04-26 article.tex
EOD

The PBD model has five biological parameters (see \ref{table:parameters}), 
which we explore in a factorial fashion, excluding some combinations. 
We assume that the speciation initiation and extinction rates
of an incipient and good species are equal ($b = b_i = b_g$ and $\mu = \mu_i = \mu_g$),
as this enables use to do more replicates 
\textbf{[NOTE: I am unconvinced. I think we should also explore $b_i \neq b_g$ and $\mu_i \neq \mu_g$]}.
We only simulate a PBD process for phylogenies in which
speciation initiation exceeds extinction rate ($b > \mu$),
and in which their difference is not too big ($b - \mu < 0.8$), 
to prevent overly taxon-poor and taxon-rich phylogenies respectively.
The parameter values chosen are a 
superset of \cite{etienne2014}, as these parameters result in reasonably
sized phylogenies and allows us to compare results. 
For the speciation initiation rate $b$, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit.
The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
events per (incipient species) lineage per time unit. For 
$\lambda = \infty$ (where we assume that in this context $10^9 \approx \infty$), 
the PBD model equals a BD model, which allows us to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit.

@@

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2 \cite{beast2} using the 
pirouette (\cite{pirouette}) package. For our site model, we assume a Jukes-Cantor 
nucleotide substitution model, as we used that in the simulation of the alignment.
For our clock model, we assume a strict clock with the same fixed rate as 
used in the simulation of the alignment \textbf{[NOTE: \cite{moller2018} did not use a
fixed clock rate, I do not see why]}. The tree prior assumed is the BD model, 
as this simplification is the goal of this research. 
Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages used have the same resolution (of $10^{-3}$ time 
units) as the alignment. The MCMC chain is run to generate 1111 states,
of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
1000 MCMC states \textbf{[NOTE: Why 1000? Why not 250? I would say 250 is preferable, as
the information will be more dense]}, the effective sample size (ESS) of the posterior must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
Would the ESS be less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

@@

\textbf{[NOTE: journal does not request for this]}
RJCB and RSE conceived the idea for this experiment and package. 
RJCB created and tested the experiment and package, 
and wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & MRCA, MDCA or random \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 20K \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set. Sampling method $M$ is random for the general
    data set. For the data set exploring the effect of sampling, MRCA is
    used for odd values of $R_i$, and MDCA is used for even values of $R_i$.
    $R_i$ is 1 for the first simulation, 2 for the next, etcetera.
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

The PBD model has five biological parameters (see \ref{table:parameters}), 
which we explore in a factorial fashion, excluding some combinations. 
We only simulate a PBD process for phylogenies in which
speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
and in which the expected number of taxa is less than 1000:

```
EQUATION SHOWING THE EXPECTED NUMBER OF TAXA < 1000
```

to prevent overly taxon-poor and taxon-rich phylogenies respectively.
The parameter values chosen are a 
superset of \cite{etienne2014}, as these parameters result in reasonably
sized phylogenies and allows us to compare results. 
For the speciation initiation rate $b$, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit.
The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
events per (incipient species) lineage per time unit. For 
$\lambda = \infty$ (where we assume that in this context $10^9 \approx \infty$), 
the PBD model equals a BD model, which allows us to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit.

@@

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2 \cite{beast2} using the 
pirouette (\cite{pirouette}) package. For our site model, we assume a Jukes-Cantor 
nucleotide substitution model, as we used that in the simulation of the alignment.
For our clock model, we assume a strict clock with the same fixed rate as 
used in the simulation of the alignment \textbf{[NOTE: \cite{moller2018} did not use a
fixed clock rate, I do not see why]}. The tree prior assumed is the BD model, 
as this simplification is the goal of this research. 
Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages used have the same resolution (of $10^{-3}$ time 
units) as the alignment. The MCMC chain is run to generate 1111 states,
of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
1000 MCMC states, the effective sample size (ESS) of the posterior must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
Would the ESS be less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

@@

\textbf{[NOTE: journal does not request for this]}
RSE conceived the idea for this experiment. 
RJCB created and tested the experiment, 
and wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b_g$                 & Speciation initiation rate of a good species & 0.1, 0.5, 1.0 \\
    $b_i$                 & Speciation initiation rate of an incipient species & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2, 0.4 \\
    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & MRCA, MDCA or random \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 20K \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set. Sampling method $M$ is random for the general
    data set. For the data set exploring the effect of sampling, MRCA is
    used for odd values of $R_i$, and MDCA is used for even values of $R_i$.
    $R_i$ is 1 for the first simulation, 2 for the next, etcetera.
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
